---
title: "dplyr"
author: "Saurabh Mishra"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to dplyr

### R Markdown

Subsetting data should be the LAST resort. Use proper data aggregation techniques or facetting in ggplot2. Or, more realistic, only subset the data as a temporary measure while you develop your elegant code for computing on or visualizing these data subsets.

Copies and excerpts of your data clutter your workspace, invite mistakes, and sow general confusion. Avoid whenever possible. Reality can also lie somewhere in between. A subset is not self-documenting and is fragile (in case of data manipulation)


### Filter 

filter() takes logical expressions and returns the rows for which all are TRUE.

filter() subsets data row-wise:

```{r cars}

library(gapminder)
library(tidyverse)

gapminder

filter(gapminder, lifeExp < 39)

filter(gapminder, country == "Canada")

filter(gapminder, country == "Rwanda", year > 1979)

filter(gapminder, country %in% c("Rwanda", "Afghanistan"))

```


### Pipe

The pipe operator takes the thing on the left-hand-side and pipes it into the function call on the right-hand-side – literally, drops it in as the first argument.

```{r pressure, echo=FALSE}

gapminder %>% head(10)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


### Select

select() takes logical expressions and returns the columns for which all are TRUE.

select() is used to subset the data on variables or columns:

```{r}

select(gapminder, year, lifeExp)

gapminder %>% select(year, lifeExp) %>% head(4)

gapminder %>%
  filter(country == "Cambodia") %>%
  select(year, lifeExp)

# R native:-
# gapminder[gapminder$country == "Cambodia", c("year", "lifeExp")]
```

select() can: 
1. rename the variables you request to keep.
2. be used with everything() to hoist a variable up to the front of the tibble.    everything() is a helper for variable selection 

```{r}

library(tidyverse)

filter(gapminder, country == "Burundi", year > 1996) %>% 
select(yr = year, lifeExp, gdpPercap) %>% 
select(gdpPercap, everything())

```


## Intermediate dplyr

### Mutate

mutate() is a function that defines and inserts new variables (columns) into a tibble. 

Imagine we wanted to recover each country’s GDP. After all, the Gapminder data has a variable for population and GDP per capita. 

```{r}
# create an explicit copy of gapminder for experimenting
(my_gap <- gapminder)

my_precious <- my_gap %>% filter(country == "Canada")

my_gap %>%
    mutate(gdp = pop * gdpPercap)

# GDP numbers are almost uselessly large and abstract. Randall Munroe of xkcd:

# "One thing that bothers me is large numbers presented without context… ‘If I added a zero to this number, would the sentence containing it mean something different to me?’ If the answer is ‘no,’ maybe the number has no business being in the sentence in the first place."


# It would be better to report GDP per capita, relative to some benchmark country. Taking Canada:

ctib <- my_gap %>%
  filter(country == "Canada")

# joining year would be a better option
my_gap <- my_gap %>%
    mutate(tmp = rep(ctib$gdpPercap, nlevels(country)),
           gdpPercapRel = gdpPercap / tmp,
           tmp = NULL) # get rid of a variable by setting it to NULL

# Canadian values for gdpPercapRel better all be 1!
my_gap %>% 
  filter(country == "Canada") %>% 
  select(country, year, gdpPercapRel)

summary(my_gap$gdpPercapRel)
# The relative GDP per capita numbers are, in general, well below 1. We see that most of the countries covered by this dataset have substantially lower GDP per capita, relative to Canada, across the entire time period.
```


### Arrange

arrange() reorders the rows in a data frame in a principled way.

```{r}

# data ordered by year then country
my_gap %>%
  arrange(year, country)

# data from 2007, sorted on life expectancy in ascending order
my_gap %>%
  filter(year == 2007) %>%
  arrange(lifeExp)

# descending order
my_gap %>%
  filter(year == 2007) %>%
  arrange(desc(lifeExp))
```


### Rename

Used to rename vairables (columns)

```{r}

my_gap %>%
    rename(life_exp = lifeExp,
           gdp_percap = gdpPercap,
           gdp_percap_rel = gdpPercapRel)

```

**Select can also be used to rename variables**


### Group By and Summarize

group_by() adds extra structure to the dataset – grouping information – which lays the groundwork for computations within the groups.

summarize() takes a dataset with n observations, computes requested summaries, and returns a dataset with 1 observation.

```{r}
my_gap %>%
  group_by(continent) %>%
  summarize(continent_obsv = n())
```

In native R, the object of class table that is returned makes downstream computation a bit fiddlier than you’d like. In this case, it’s too bad the continent levels come back only as names and not as a proper factor, with the original set of levels. This is an example of how the tidyverse smooths transitions where you want the output of step i to become the input of step i + 1.

```{r}

# tally() is a convenience function that counts rows and honors groups.
my_gap %>%
  group_by(continent) %>%
  tally()

# count() is an even more convenient function that does both grouping and counting.
my_gap %>% 
  count(continent)

```

```{r}

# number of unique countries in each continent
my_gap %>%
  group_by(continent) %>%
  summarize(n = n(),
            n_countries = n_distinct(country))

# average life expectancy by continent
my_gap %>%
  group_by(continent) %>%
  summarize(avg_lifeExp = mean(lifeExp))

#summarize_at() applies the same summary function(s) to multiple variables. Let’s compute average and median life expectancy and GDP per capita by continent by year … but only for 1952 and 2007.
my_gap %>%
  filter(year %in% c(1952, 2007)) %>%
  group_by(continent, year) %>%
  summarize_at(vars(lifeExp, gdpPercap), funs(mean, median))

# What are the minimum and maximum life expectancies seen by year in Asia
my_gap %>%
  filter(continent == "Asia") %>%
  group_by(year) %>%
  summarize(min_lifeExp = min(lifeExp), max_lifeExp = max(lifeExp))
```

